---
title: "Grayling_FinalFits"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library("hypergeo")
library(stats)
```

This code takes data from various experimental conditions and fits the eDNA and eRNA decay model for the shedding and decay rates.
This model fits individual shedding rates (S) and decay rate (k) for each animal within an experiment.
NOTE: In all cases, we fit the log of S and k rather than S and k directly, in order to force these values to stay positive.
## Experiment types:
  eDNA shedding and decay at different flow rates. This is fit using maximum likelihood because of the discontinuity of the function.
  eDNA and eRNA shedding and decay in a static tank. These are fit using a non-linear least squares fitting.

# DIFFERENT FLOW RATES

Load data from an experiment with different flow rates
Variables: Fish (e.g. 1-4), Time (Hours) (i.e. when the eDNA was sampled), Copies/L (can be changed to just Copies) 

## Define the model. 
In this case, we define separate sections of the output where the flow rate is constant.

Variables to alter:
  tl = duration of each flow rate [example = 120]min
  v = volume of water [example = 285]L
  uvec = each flow rate used in the experiment [example = c(1,5,10,20)/v]L/min
  s <- ps[1:4] = shedding rate parameters for the # of animals in experiment [example = 4]
  k <- pk[1:4] = decay rate for each fish

We then define the starting concentration (C0) at each segment according to the end concentration at the previous flow rate.

```{r flow rate definitions}

fr_data <- read.csv(paste("eDNA_Data.xlsx - FlowRate Data.csv",sep=""), header=TRUE)

# Define model
fr_model <- function(t,s,k,c0emp){
    # This is the amount of time between when the velocities were changes
    tl <- 2 # Hours
    # This is the total volume
    v <- 285 # L
    # These are the empirical velocities, changed into flow rates by dividing by volume
    uvec <- c(1,5,10,20)/v*60 # To convert to hours
    # Take the exponent to ensure these stay positive, since negatives don't make sense biologically
    es <- exp(s)
    ek <- exp(k)
    # Define concentration c0 and constant term in solution c0'
    c0 <- c(0,0,0,0)
    c0p <- c(0,0,0,0)
    # Loop to define these
    for (i in c(0,1,2,3)){
        # Note: Set c0 to empirical value at 0
        c0[i+1] <- ifelse(i==0, c0emp, exp(-(ek+uvec[i])*tl)*c0p[i] + es/v/(ek+uvec[i]))
        # This is the value at the start of each segment of constant velocity
        c0p[i+1] <- ifelse(i==0, c0[1] - es/v/(ek+uvec[1]), c0[i+1] - es/v/(ek+uvec[i+1]))
    }
    # Now get return variable. The ifelse statement find which segment of the solution we are on.
    # In other words, what the velocity of the water is.
    ct <- ifelse(t<=tl,exp(-(ek+uvec[1])*t)*c0p[1]+es/v/(ek+uvec[1]),
                ifelse(t<=2*tl,exp(-(ek+uvec[2])*(t-tl))*c0p[2]+es/v/(ek+uvec[2]),
                ifelse(t<=3*tl,exp(-(ek+uvec[3])*(t-2*tl))*c0p[3]+es/v/(ek+uvec[3]),
                       exp(-(ek+uvec[4])*(t-3*tl))*c0p[4]+es/v/(ek+uvec[4]))))
}

# Define the log likelihood function.
ll_fr <- function(pa,dft){
    # pa are parameters, df is data for a single fish
    # Shed rates
    s <- pa[1] # One for each fish
    # Decay rate
    k <- pa[2] # One for each fish
    times <- dft[dft$Hours>0,'Hours']
    vals <- dft[dft$Hours>0,'Copies1Lsc']
    c0 <- dft[dft$Hours==0,'Copies1Lsc']
    mod <- fr_model(times,s,k,c0)
    logpmf <- -mod + vals*log(mod)-lgamma(vals+1)
    ll <- -sum(logpmf)
    return(ll)
}

# Now use optimize to get the optimal values for the shed rates and the decay rate, and print them out
# The par values are just initial guesses for the values of the shed rates and decay rates
# but note they are the log of the actual values since we transform them by taken the exponent in the fr_model function
df <- split(fr_data,fr_data$Organism)
# Make storage for shed and decay rates
ps <- rep(0,length(df))
pk <- rep(0,length(df))
for (i in 1:length(df)){
  dft<-df[[i]]
  fr_opt <- optim(par=c(11,-23),ll_fr,dft=dft)
  ps[[i]]<-fr_opt$par[1]
  pk[[i]]<-fr_opt$par[2]
}
print("Optimization results from flow rate experiment")
exp(ps)
exp(pk)
# Save to files
write.csv(data.frame(s=exp(ps),k=exp(pk)),"flow_rate_fits.csv", row.names = FALSE)
```
# STATIC TANK EXPERIMENT
In this experiment, a fish was introduced in the tank, and then removed after a certain time.

## Define the necessary functions
  t_sc = time the animal was removed from the environment (230 min here)
  s <- shedding rate
  k <- decay rate
  
Define the model for the copies
```{r define static tank}
# Read in data
sc_data <- read.csv(paste("eDNA_Data.xlsx - Static.csv"))
# Separate by fish
sc_individual <- split(sc_data,sc_data$BioRep)

t_sc <- 230/60 #in hours
sc_model <- function(t,s,k,c0emp,pv){
    # Exponent of s and k to force positive
    es <- exp(s)
    ek <- exp(k)
    # This term depends on what the time is. After the fish is removed at t=t_sc, the shed rate is 0.
    # The form of the function here is fairly complicated because of the volume function, for which the parameters
    # given by pv. 
    hypt1 <- ifelse(t<=t_sc,exp((pv[2]+ek)*t)*hypergeo(1,1+ek/pv[2],2+ek/pv[2],-pv[3]*exp(pv[2]*t)/pv[1]),
                exp((pv[2]+ek)*t_sc)*hypergeo(1,1+ek/pv[2],2+ek/pv[2],-pv[3]*exp(pv[2]*t_sc)/pv[1]))
    hypt2 <- hypergeo(1,1+ek/pv[2],2+ek/pv[2],-pv[3]/pv[1])
    hypt <- Re((hypt1-hypt2)/(pv[1]*(pv[2]+ek)))
    return(exp(-ek*t)*(c0emp+hypt*es))
}

#Now we have to get the function for how the volume decays

# Get all coefficients for parameters, one for each Organism
pvs <- Map(nls,sc_individual, MoreArgs=list(formula = Volume ~ p1*exp(-p2 * Hours) + p3, start=list(p1=150,p2=0.0005*60,p3=350))) # p2 has time units, so 0.03 for hours, but 0.0005 for minutes
# Save to file
write.csv(t(data.frame(coef(pvs[[1]]),coef(pvs[[2]]),coef(pvs[[3]]),coef(pvs[[4]]))),"static_tank_fits_volume.csv",row.names = FALSE)
```

```{r static tank eDNA}

sdna <- c(0,0,0,0)
kdna <- c(0,0,0,0)
sdna_err <- c(0,0,0,0)
kdna_err <- c(0,0,0,0)

# Loop over each fish
for (i in 1:length(sc_individual)){
  # Temporary data to fit
  dft <- sc_individual[[i]]
  # Remove NA
  dft <- dft[complete.cases(dft$eDNA1Lsc),]
  times <- dft[dft$Hours>0,'Hours']
  vals <- dft[dft$Hours>0,'eDNA1Lsc']
  c0 <- mean(dft[dft$Hours==0,'eDNA1Lsc'])
  # Define the function to fit within the loop so that it has the appropriate values
  sc_nls_fit <- function(s,k){
    # s is shed rate
    # k is decay rate
    mod <- sc_model(times,s,k,c0,coef(pvs[[i]]))
    return(mod)
  }
  # Use nls to fit. Technically this uses gaussian errors rather than Poisson
  fit <- nls(vals ~ sc_nls_fit(s,k),start=list(s=19,k=-3))
  # Save to file
  sdna[i] <- exp(coef(fit))[[1]]
  sdna_err[i] <- exp(coef(fit)[[1]])*sqrt(vcov(fit))[1,1]
  kdna[i] <- exp(coef(fit))[[2]]
  kdna_err[i] <- exp(coef(fit)[[2]])*sqrt(vcov(fit))[2,2]
  # Print out
  print("S and standard error")
  print(sdna[i])
  print(sdna_err[i])
  print("k standard error")
  print(kdna[i])
  print(kdna_err[i])
}

# Save to file
write.csv(data.frame(s=sdna,serr=sdna_err,k=kdna,kerr=kdna_err),"static_tank_fits_eDNA.csv",row.names=FALSE)
```

```{r static tank eRNA}
srna <- c(0,0,0)
krna <- c(0,0,0)
srna_err <- c(0,0,0)
krna_err <- c(0,0,0)

for (i in 1:3){
  # Temporary data to fit
  dft <- sc_individual[[i]]
  # Remove NA
  dft <- dft[complete.cases(dft$eRNA1Lsc),]
  times <- dft[dft$Hours>0,'Hours']
  vals <- dft[dft$Hours>0,'eRNA1Lsc']
  c0 <- mean(dft[dft$Hours==0,'eRNA1Lsc'])
  # Define the function to fit
  sc_nls_fit <- function(s,k){
    # s is shed rate
    # k is decay rate
    mod <- sc_model(times,s,k,c0,coef(pvs[[i]]))
    return(mod)
  }
  fit <- nls(vals ~ sc_nls_fit(s,k),start=list(s=19,k=-3))
  # Save to file
  srna[i] <- exp(coef(fit))[[1]]
  srna_err[i] <- exp(coef(fit)[[1]])*sqrt(vcov(fit))[1,1]
  krna[i] <- exp(coef(fit))[[2]]
  krna_err[i] <- exp(coef(fit)[[2]])*sqrt(vcov(fit))[2,2]
  # Print out
  print("S and standard error")
  print(srna[i])
  print(srna_err[i])
  print("k standard error")
  print(krna[i])
  print(krna_err[i])
}
# Save to file
write.csv(data.frame(s=srna,serr=srna_err,k=krna,kerr=krna_err),"static_tank_fits_eRNA.csv",row.names=FALSE)
```

```{r summary}
sdna_sum <- c(exp(ps),sdna)
print("Mean and standard deviation of S")
mean(sdna_sum)
sd(sdna_sum)
print("Mean and standard deviation of k")
mean(kdna)
sd(kdna)
print("Mean and standard deviaiton of S (eRNA)")
mean(srna)
sd(srna)
print("Mean and standard deviaiton of k (eRNA)")
mean(krna)
sd(krna)
```